Modeling the uncertainties of solar-system ephemerides for robust gravitational-wave searches with pulsar timing arrays

### 2020-02-21 paper-apj/main.tex
Changes for ApJ resubmit

%% Mark off the abstract in the ``abstract'' environment. 
\begin{abstract}
The regularity of pulsar emissions becomes apparent once we reference the pulses' times of arrivals to the inertial rest frame of the solar system. It follows that errors in the determination of Earth's position with respect to the solar-system barycenter can appear as a time-correlated bias in pulsar-timing residual time series, affecting the searches for low-frequency gravitational waves performed with pulsar timing arrays.
Indeed, recent array datasets yield different gravitational-wave background upper limits and detection statistics when analyzed with different solar-system ephemerides. Crucially, the ephemerides do not generally provide usable error representations.
In this article we describe the motivation, construction, and application of a physical model of solar-system ephemeris uncertainties, which focuses on the degrees of freedom (Jupiter's orbital elements) most relevant to gravitational-wave searches with pulsar timing arrays.
This model, \textsc{BayesEphem}, was used to derive ephemeris-robust results in NANOGrav's 11-yr stochastic-background search, and it provides a foundation for future searches by NANOGrav and other consortia.
The analysis and simulations reported here suggest that ephemeris modeling reduces the gravitational-wave sensitivity of the 11-yr dataset; and that this degeneracy will vanish with improved ephemerides and with the longer pulsar timing datasets that will become available in the near future.
\end{abstract}

@@

% What is our place in the Universe? While that question may lie in the philosophical realm for most, it represents a very practical concern for pulsar astronomers
Pulsar timing exploits the remarkable regularity of millisecond-pulsar emissions to extract accurate system parameters from time-of-arrival (TOA) datasets \citep{2012hpa..book.....L},
by fitting precise \emph{timing models} that account for all pulse delays and advances, from generation near the neutron stars to detection at the radiotelescopes \citep{2013CQGra..30v4001L}.
% Within the precise timing model used to extract physical parameters from pulse time-of-arrival (TOA) datasets, the largest contribution is
The largest time-dependent term in the model is the \emph{R{\o}mer delay} \citep{Roemer1676} due to the motion of Earth around the solar-system barycenter (SSB), with magnitude $\sim 500$ s. Solar-system ephemerides (SSEs), such as those produced by the Jet Propulsion Laboratory (JPL; see \citealt{2009IPNPR.178C...1F,2014IPNPR.196C...1F,de435,de436,de438}), are used to convert observatory TOAs to the notional coordinate time of an inertial frame centered at the SSB.
It follows that errors in our estimate of Earth's trajectory around the SSB produce a time-dependent bias in the TOAs.

Throughout many years of pulsar-timing studies and discoveries, SSE errors were always considered negligible compared to all other sources of noise and uncertainty \citep{1990ApJ...361..300F,2006MNRAS.372.1549E}.
% MV: These are Foster and Backer 1990 (what do they say exactly?) and the tempo2 paper (does it have a budget?)
TOAs are now being collected with ever greater time spans and timing precisions, especially so by the \emph{pulsar-timing-array} (PTA) collaborations seeking to detect gravitational waves (GWs) as correlated residuals (i.e., TOAs minus timing model) in multi-pulsar datasets \citep{saz78,det79,fb90,ml13,dcl+16,h13,v+16}.
The estimated magnitude of the GW signature is $\sim 100$ ns or less, leading to the recent suggestion that SSE errors could measurably bias GW results obtained from these datasets \citep{2016MNRAS.455.4339T,2019ApJ...876...55R}.
% maybe: 2013AdSpR..52.1602D - Interplanetary spacecraft navigation using pulsars
%
\begin{figure*}[t]
\begin{center}
    \includegraphics[width=0.85\textwidth]{figures/GWposterior.pdf}
\end{center}
\vspace{-12pt}
    \caption{(\textbf{left}) Bayesian posteriors for the amplitude $A_\mathrm{GWB}$ (at $f = \mathrm{yr}^{-1}$) of the GW stochastic background, modeled as a fixed power law with slope $\gamma = 11/3$, as appropriate for a population of inspiraling and GW-emitting supermassive black-hole binaries.
    The posteriors are computed for the NANOGrav 11-year dataset using individual JPL ephemerides (dotted lines), and \textsc{BayesEphem} (solid lines). The GW model (``model 2A'' in \citealt{2018ApJ...859...47A}) does not include Hellings--Downs correlations, which however modify results only marginally.
    The incorporation of explicit SSE uncertainties into the analysis via \textsc{BayesEphem} leads to substantially lower evidence of a GW background, and to much greater consistency among the different SSEs.
    % This plot demonstrates that \textsc{BayesEphem} bridges the JPL SSEs successfully; in doing so, it also removes most evidence for the presence of a GW background.
    (\textbf{right}) Same posteriors as in the left panel, shown with a logarithmic vertical scale that can be mapped to approximate GW vs.\ noise-only Bayes factors, by way of the Savage--Dickey formula $P(A_\mathrm{GWB}=0)/P(A_\mathrm{GW}\neq 0) = p(A_\mathrm{GWB}=0|\mathrm{data})/p_\mathrm{prior}(A_\mathrm{GWB}=0)$ \citep{d71}.
    The application of \textsc{BayesEphem} brings all Bayes factors close to unity, indicating no evidence for GWs once SSE uncertainties are taken into consideration.
    Bayes factors and 95\% $A_\mathrm{GWB}$  upper limits calculated from these curves are listed in Tables \ref{tab:bf} and \ref{tab:upper}.
    % We rescale the vertical axis of the logarithmic plot on the right to show these factors. 
    \label{fig:posteriors}}
\end{figure*}

Within the North American Nanohertz Observatory for Gravitational Waves (NANOGrav), we encountered this bias firsthand in early 2017, when we observed that switching between the more recent SSEs issued by JPL (specifically DE421, DE430, DE435, and DE436: \citealt{2009IPNPR.178C...1F,2014IPNPR.196C...1F,de435,de436}) led to discrepant results (GW upper limits and detection statistics) in the search for a stochastic signal from the population of supermassive black-hole binaries, formed in the centers of galaxies following major-merger events, as performed on the NANOGrav 11-year dataset \citep{2018ApJS..235...37A,2018ApJ...859...47A}.
See Fig.\ \ref{fig:posteriors} for the Bayesian posterior probabilities of the GW-background (GWB) amplitude, as obtained with a range of SSEs. We verified that other recent PTA results \citep{2015Sci...349.1522S,2016ApJ...821...13A} would be similarly affected.

@@

Across these ephemerides, the orbit of Earth relative to the Sun is consistent at the $3$-m ($\sim 10$-ns) level, after applying an overall rotation with respect to the International Celestial Reference Frame (within the uncertainties of that ``tie''), and a rotation \emph{rate} about the ecliptic pole\footnote{\label{footnote:rate}This accounts for differences in the estimated semi-major axis of the Earth-Moon--barycenter orbit, which gives rise to a linear rate in estimated ecliptic longitude.}.
However, the orbit of the Sun and therefore the orbit of Earth (both relative to the SSB) match only at the $100$-m ($\sim 300$-ns) level across ephemerides (see Fig.\ \ref{fig:earthsun}). This discrepancy arises from the estimated positions of Jupiter, Saturn, Uranus, and Neptune. Using a simple dynamical model of the SS (i.e., integrating the equations of Newtonian gravity for the eight planets and the Sun), it is easy to show that if we perturb the masses or the orbits of the outer planets, we affect the Sun-to-SSB and Earth-to-SSB trajectories through the resulting redefinition of the SSB, rather than through the very minor changes in the gravitational pull of the outer planets (see also \citealt{2019MNRAS.489.5573G}).
%
\begin{figure}[t]
    \centering
    \includegraphics[width=0.49\columnwidth]{figures/EarthSunDiff.pdf}
    \includegraphics[width=0.49\columnwidth]{figures/SunSSBDiff.pdf}
    \caption{Differences in Earth $-$ Sun position, $\mathbf{r}^{(3)} - \mathbf{r}^\mathrm{Sun}$, and in Sun $-$ SSB position, $\mathbf{r}^\mathrm{Sun} - \mathbf{r}^\mathrm{SSB}$, between pairs of JPL SSEs. Note the difference in the vertical scale of the two panels.
    Earth $-$ SSB plots, not included here, would appear essentially the same as Sun $-$ SSB plots, since the Earth $-$ Sun differences are very small. For reference, 2.5~m $\approx$ 8~ns, and 0.1~km $\approx$ 300~ns, bracketing the range of amplitudes expected for GW signals in PTA datasets.
    We plot SSE equatorial-coordinate differences over the approximate span of NANOGrav's 11-yr dataset, and we apply the best-fit frame rotation that minimizes the 3D norm of the difference. Dotted curves are zoomed $\times 10$ vertically. Dotted vertical lines mark the end of the DE421 and DE430 fitted datasets.
    For clarity, we also remove a $\sim 100$-m mean in each coordinate in the Sun--SSB differences; such a constant offset does not affect PTA likelihoods.
    \label{fig:earthsun}}
\end{figure}

@@

The search for stochastic GW signals with PTAs exploits the distinctive quadrupolar signature in the inter-pulsar correlations of timing-model residuals \citep{hd83}.
Residuals are obtained after applying a chain of corrections that convert the TOA measured at the radiotelescope to the notional emission time in the pulsar system (see, e.g., \citealt{ehm06}):
%
\begin{equation}
    t_\mathrm{e}^\mathrm{psr} = t_\mathrm{a}^\mathrm{obs} - \Delta_\odot - \Delta_\mathrm{IS} - \Delta_\mathrm{B},
\end{equation}
%
where $t_\mathrm{e}^\mathrm{psr}$ and $t_\mathrm{a}^\mathrm{obs}$ are the emission and arrival times, $\Delta_\odot$ captures corrections between the observatory and SSB frames, $\Delta_\mathrm{IS}$ describes corrections between the SSB frame and the pulsar-system barycenter, and $\Delta_\mathrm{B}$ models corrections from the pulsar system barycenter to the pulsar frame (which are relevant for binary systems).
Among these terms, $\Delta_\mathrm{IS}$ is very large, but changes very slowly over the duration of pulsar datasets, and thus maps to a constant phase offset. Next comes the R{\o}mer delay \citep{Roemer1676}, corresponding to the light travel time ($\sim 500$ s) between the observatory at $\mathbf{r}^\mathrm{obs}$ and the SSB at $\mathbf{r}^\mathrm{SSB}$:
% MV: the sign here is consistent with the R{\o}mer delay being an actual delay (the pulse takes extra time traveling from SSB-impinging wavefront to observatory)
%
\begin{equation}
\label{eq:Roemer}
\Delta_{\mathrm{R}\odot} =
t_\mathrm{a}^\mathrm{obs} - t_\mathrm{a}^\mathrm{SSB} =
-\frac{\bigl(\mathbf{r}^\mathrm{obs}(t_\mathrm{a}^\mathrm{obs}) - \mathbf{r}^\mathrm{SSB}(t_\mathrm{a}^\mathrm{obs})\bigr) \cdot \hat{\mathbf{p}}}{c},
\end{equation}
%
where $\hat{\mathbf{p}}$ is the unit vector in the direction of the pulsar. In fact, $\hat{\mathbf{p}}$ is determined from pulsar-timing data mainly through the time dependence of Eq.\ \eqref{eq:Roemer}.

@@

\citet{2010ApJ...720L.201C} discuss pulsar-timing's potential to constrain the masses of outer planets. The uncertainties in current IAU best estimates (derived from spacecraft tracking and Doppler studies: \citealt{iaumasses,jh+2000,2006AJ....132.2520J,2014AJ....148...76J,2009AJ....137.4322J}) give rise to R{\o}mer corrections comparable to the orbit errors. To wit, these scale as $(\delta m^\mathrm{planet}/m^\mathrm{SS}) \times \mathbf{r}$, so
%
\begin{equation}
    \begin{aligned}
    \delta x^\mathrm{SSB}(\delta m^{(5)}) & \approx (1.6 \!\times\! 10^{-11})\!\times\!
    % (7.8 \!\times\! 10^{8} \, \mathrm{km})
    (5.2 \, \mathrm{AU})
    \approx 12 \, \mathrm{m} \approx 40 \, \mathrm{ns}, \\
    \delta x^\mathrm{SSB}(\delta m^{(6)}) & \approx (8.2 \!\times\! 10^{-12}) \!\times\!
    % (1.4 \!\times\! 10^{9} \, \mathrm{km})
    (9.4 \, \mathrm{AU})
    \approx 12 \, \mathrm{m} \approx 40 \, \mathrm{ns}, \\
    \delta x^\mathrm{SSB}(\delta m^{(7)}) & \approx (3.2 \!\times\! 10^{-11}) \!\times\!
    % (2.9 \!\times\! 10^{9} \, \mathrm{km})
    (19 \, \mathrm{AU})
    \approx 90 \, \mathrm{m} \approx 300 \, \mathrm{ns}, \\
    \delta x^\mathrm{SSB}(\delta m^{(8)}) & \approx (8.0 \!\times\! 10^{-11}) \!\times\!
    % (4.5 \!\times\! 10^{9} \, \mathrm{km})
    (30 \, \mathrm{AU})
    \approx 360 \, \mathrm{m} \approx 1.2 \, \mu\mathrm{s}.
    % (1200 \, \mathrm{ns}).
    \end{aligned}
\end{equation}

These corrections yield delays with the same periods as the corresponding planets, so again they may be observable in PTA datasets for Jupiter and Saturn, but not for Uranus and Neptune.
% MV: say anything about the decay of accuracy past the end of the data?
%
\begin{figure*}[t]
    \centering
    \includegraphics[width=\columnwidth]{figures/roemerdiff.pdf}
    \caption{Differences in R{\o}mer delays computed with DE421/DE430/DE435/DE438 and taking DE436 as reference for 100 simulated pulsars randomly distributed across the sky, plotted after subtracting the best-fitting quadratic and yearly sinusoid. The darker curves with various dashing styles allow a comparison of the same pulsars across the three differences. The plot covers the span of the NANOGrav 11-yr dataset.
    For any given pulsar, these differences are reduced to less than 10 ns after subtracting the best-fitting linear combination of \textsc{BayesEphem} partials, dotted into the pulsar position to obtain R{\o}mer delays.
    }
    \label{fig:Roemer}
\end{figure*}

@@

Note that we do not expect this term to affect GW posteriors: both static and uniform rotations of the SSE frame are absorbed in the estimated positions and proper motions of the pulsars. We omit the former altogether, and include $\omega^{\hat{z}}$ as a check.
%
\item \textbf{TOA delays generated by perturbing Jupiter's average orbital elements}, as given by
\begin{equation}
\label{eq:orbitperturb}
    \delta t_{\mathrm{R}\odot}(\delta a^\mu) = -\delta a^\mu
    \frac{m^{(5)}}{m^\mathrm{SS}} \frac{\partial \mathbf{x}^{(5)}}{\partial a^\mu} \cdot \frac{\mathbf{p}}{c},
\end{equation}
%
where the six $a^\mu$ are the six J2000 Keplerian elements (semimajor axis, eccentricity, inclination, mean longitude, longitude of the perihelion, longitude of the ascending node), and the $\delta a^\mu$ are their perturbations. Alternatively, we can formulate the perturbations in terms of Brouwer and Clemence's ``Set III'' parameters \citep{1961mcm..book.....B}, for which we have access to uncertainty estimates for certain JPL SSEs (\citealt{de434,de438}; see Fig.\ \ref{fig:setIIIposteriors}). The two formulations are largely equivalent with respect to their effect on GW searches.
% quasi-Keplerian parameters
% a: semimajor axis (in AU)
% e: eccentricity (in rad? should be dimensionless)
% I: inclination (in deg)
% L: mean longitude (in deg)
% longitude of the periapsis (in deg)
% longitude of the ascending node (in deg)
% rates in these units per century -> one of the rates corresponds to the period

To implement the R{\o}mer-delay perturbations, we begin with approximate values for the $a^\mu$ and their rates of change.\footnote{The six $a^\mu$ plus the rate of change $\dot{a}^4 \equiv \dot{l}$ of mean longitude specify Jupiter's orbit as the osculating ellipse at the J2000 reference epoch (MJD 2451545); the remaining five rates encode the secular evolution of Jupiter's orbit due to SSB bodies other than the Sun, and to other physical effects. See \href{https://ssd.jpl.nasa.gov/txt/aprx_pos_planets.pdf}{https://ssd.jpl.nasa.gov/txt/aprx\_pos\_planets.pdf}.}
%\citep{ssdelements}.
We vary these $a^\mu$ to minimize the (root-mean-square) difference between our quasi-Keplerian orbits and DE436, integrated between years 2000 and 2020. The resulting orbits are within 1\% of DE436, which ensures similar accuracy for the orbit partials, more than enough for our purposes.
%
\begin{figure*}[t]
    \centering
    \includegraphics[width=\columnwidth]{figures/JupiterPartials.pdf}
    \includegraphics[width=\columnwidth]{figures/JupiterPartialsSVD.pdf}
    \caption{\textbf{Top}: perturbative partials ($x$, $y$, and $z$ coordinates, geometrized units of seconds) for Jupiter's orbit between the years 2000 and 2020, as obtained by varying the six Keplerian elements in a quasi-Keplerian model that includes also their rates. The baseline elements and rates are adjusted to fit DE436 orbits.
    Here $a$ is the semimajor axis of the orbit, $e$ the eccentricity, $\imath$ the inclination, $\ell$ the mean longitude, $\varpi$ the longitude of the periapsis, and $\Omega$ the longitude of the ascending node. 
    \textbf{Bottom}: singular-value-decomposition vector time series obtained from the partials in the top panel.}
    \label{fig:jupiterorbit}
\end{figure*}

@@

We do not include Uranus and Neptune's orbital perturbations, which lead to $\delta t_{\mathrm{R\odot}}$ with linear time dependencies that are absorbed entirely by timing-model parameters. By contrast, we repeat the procedure that we just outlined for Saturn, but find that GW posteriors are barely affected for orbital perturbations of magnitude comparable to the differences among the JPL SSEs. Thus in our use of \textsc{BayesEphem} we usually omit Saturn orbit perturbations. These may prove more important as datasets increase in length.
\end{itemize}

>>>>

%% Mark off the abstract in the ``abstract'' environment. 
\begin{abstract}
The regularity of pulsar emissions becomes apparent once we reference the pulses' times of arrivals to the inertial rest frame of the solar system. It follows that errors in the determination of Earth's position with respect to the solar-system barycenter can appear as a time-correlated bias in pulsar-timing residual time series, affecting the searches for low-frequency gravitational waves performed with pulsar timing arrays.
Indeed, recent array datasets yield different gravitational-wave background upper limits and detection statistics when analyzed with different solar-system ephemerides. Crucially, the ephemerides do not generally provide usable error representations.
In this article we describe the motivation, construction, and application of a physical model of solar-system ephemeris uncertainties, which focuses on the degrees of freedom (Jupiter's orbital elements) most relevant to gravitational-wave searches with pulsar timing arrays.
This model, \textsc{BayesEphem}, was used to derive ephemeris-robust results in NANOGrav's 11-yr stochastic-background search, and it provides a foundation for future searches by NANOGrav and other consortia.
The analysis and simulations reported here suggest that ephemeris modeling reduces the gravitational-wave sensitivity of the 11-yr dataset; and that this degeneracy will vanish with improved ephemerides and with pulsar-timing datasets that extend well beyond a single Jovian orbital period.
\end{abstract}

@@

% What is our place in the Universe? While that question may lie in the philosophical realm for most, it represents a very practical concern for pulsar astronomers
Pulsar timing exploits the remarkable regularity of millisecond-pulsar emissions to extract accurate system parameters from time-of-arrival (TOA) datasets \citep{2012hpa..book.....L},
by fitting precise \emph{timing models} that account for all pulse delays and advances, from generation near the neutron stars to detection at the radio telescopes \citep{2013CQGra..30v4001L}.
% Within the precise timing model used to extract physical parameters from pulse time-of-arrival (TOA) datasets, the largest contribution is
The largest time-dependent term in the model is the \emph{R{\o}mer delay} \citep{Roemer1676} due to the motion of Earth around the solar-system barycenter (SSB), with magnitude $\sim 500$ s. Solar-system ephemerides (SSEs), such as those produced by the Jet Propulsion Laboratory (JPL; see \citealt{2009IPNPR.178C...1F,2014IPNPR.196C...1F,de435,de436,de438}), are used to convert observatory TOAs to the notional coordinate time of an inertial frame centered at the SSB.
It follows that errors in our estimate of Earth's trajectory around the SSB produce a time-dependent bias in the TOAs.

Throughout many years of pulsar-timing studies and discoveries, SSE errors were always considered negligible compared to all other sources of noise and uncertainty \citep{1990ApJ...361..300F,2006MNRAS.372.1549E}.
% MV: These are Foster and Backer 1990 (what do they say exactly?) and the tempo2 paper (does it have a budget?)
TOAs are now being collected with ever greater time spans and timing precisions, especially so by the \emph{pulsar-timing-array} (PTA) collaborations seeking to detect gravitational waves (GWs) as correlated residuals (i.e., TOAs minus timing model) in multi-pulsar datasets \citep{saz78,det79,fb90,ml13,dcl+16,h13,v+16}.
The estimated magnitude of the GW signature is $\sim 100$ ns or less, leading to the recent suggestion that SSE errors could measurably bias GW results obtained from these datasets \citep{2016MNRAS.455.4339T}. Indeed, while SSE corrections and GWs have different spatial-correlation signatures across multiple pulsars, the two may yet be degenerate when sampled with a limited number of PTA pulsars \citep{2019ApJ...876...55R}.
% maybe: 2013AdSpR..52.1602D - Interplanetary spacecraft navigation using pulsars
%
\begin{figure*}[t]
\begin{center}
    \includegraphics[width=0.85\textwidth]{GWposterior.pdf}
\end{center}
\vspace{-12pt}
    \caption{(\textbf{left}) Bayesian posteriors for the amplitude $A_\mathrm{GWB}$ (at $f = \mathrm{yr}^{-1}$) of the GW stochastic background, modeled as a fixed power law with slope $\gamma = 11/3$, as appropriate for a population of inspiraling and GW-emitting supermassive black-hole binaries.
    The posteriors are computed for the NANOGrav 11-year dataset using individual JPL ephemerides (dotted lines), and \textsc{BayesEphem} (solid lines). The GW model (``model 2A'' in \citealt{2018ApJ...859...47A}) does not include Hellings--Downs correlations, which however modify results only marginally.
    The incorporation of explicit SSE uncertainties into the analysis via \textsc{BayesEphem} leads to substantially lower evidence of a GW background, and to much greater consistency among the different SSEs.
    % This plot demonstrates that \textsc{BayesEphem} bridges the JPL SSEs successfully; in doing so, it also removes most evidence for the presence of a GW background.
    (\textbf{right}) Same posteriors as in the left panel, shown with a logarithmic vertical scale that can be mapped to approximate GW vs.\ noise-only Bayes factors, by way of the Savage--Dickey formula $P(A_\mathrm{GWB}=0)/P(A_\mathrm{GW}\neq 0) = p(A_\mathrm{GWB}=0|\mathrm{data})/p_\mathrm{prior}(A_\mathrm{GWB}=0)$ \citep{d71}.
    The application of \textsc{BayesEphem} brings all Bayes factors close to unity, indicating no evidence for GWs once SSE uncertainties are taken into consideration.
    Bayes factors and 95\% $A_\mathrm{GWB}$  upper limits calculated from these curves are listed in Tables \ref{tab:bf} and \ref{tab:upper}.
    % We rescale the vertical axis of the logarithmic plot on the right to show these factors. 
    \label{fig:posteriors}}
\end{figure*}

Within the North American Nanohertz Observatory for Gravitational Waves (NANOGrav), we encountered this bias first hand in early 2017, when we observed that switching between the more recent SSEs issued by JPL (specifically DE421, DE430, DE435, and DE436: \citealt{2009IPNPR.178C...1F,2014IPNPR.196C...1F,de435,de436}) led to discrepant results (GW upper limits and detection statistics) in the search for a stochastic signal from the population of supermassive black-hole binaries, formed in the centers of galaxies following major-merger events, as performed on the NANOGrav 11-year dataset \citep{2018ApJS..235...37A,2018ApJ...859...47A}.
See Fig.\ \ref{fig:posteriors} for the Bayesian posterior probabilities of the GW-background (GWB) amplitude, as obtained with a range of SSEs. We verified that other recent PTA results \citep{2015Sci...349.1522S,2016ApJ...821...13A} would be similarly affected.

@@

Across these ephemerides, the orbit of Earth relative to the Sun is consistent at the $3$-m ($\sim 10$-ns) level, after applying an overall rotation with respect to the International Celestial Reference Frame (within the uncertainties of that ``tie''), and a rotation \emph{rate} about the ecliptic pole\footnote{\label{footnote:rate}This accounts for differences in the estimated semi-major axis of the Earth-Moon--barycenter orbit, which gives rise to a linear rate in estimated ecliptic longitude.}.
However, the orbit of the Sun and therefore the orbit of Earth (both relative to the SSB) match only at the $100$-m ($\sim 300$-ns) level across ephemerides (see Fig.\ \ref{fig:earthsun}). This discrepancy arises from the estimated positions of Jupiter, Saturn, Uranus, and Neptune. Using a simple dynamical model of the SS (i.e., integrating the equations of Newtonian gravity for the eight planets and the Sun), it is easy to show that if we perturb the masses or the orbits of the outer planets, we affect the Sun-to-SSB and Earth-to-SSB trajectories through the resulting redefinition of the SSB, rather than through the very minor changes in the gravitational pull of the outer planets (see also \citealt{2019MNRAS.489.5573G}).
%
\begin{figure}[t]
    \centering
    \includegraphics[width=0.49\columnwidth]{EarthSunDiff.pdf}
    \includegraphics[width=0.49\columnwidth]{SunSSBDiff.pdf}
    \caption{Differences in Earth $-$ Sun position, $\mathbf{r}^{(3)} - \mathbf{r}^\mathrm{Sun}$, and in Sun $-$ SSB position, $\mathbf{r}^\mathrm{Sun} - \mathbf{r}^\mathrm{SSB}$, between pairs of JPL SSEs. Note the difference in the vertical scale of the two panels.
    Earth $-$ SSB plots, not included here, would appear essentially the same as Sun $-$ SSB plots, since the Earth $-$ Sun differences are very small. For reference, 2.5~m $\approx$ 8~ns, and 0.1~km $\approx$ 300~ns, bracketing the range of amplitudes expected for GW signals in PTA datasets.
    We plot SSE equatorial-coordinate differences over the approximate span of NANOGrav's 11-yr dataset, and we apply the best-fit frame rotation that minimizes the 3D norm of the difference. Dotted curves are zoomed $\times 10$ vertically. Dotted vertical lines mark the end of the DE421 and DE430 fitted datasets.
    For clarity, we also remove a $\sim 100$-m mean in each coordinate in the Sun--SSB differences; such a constant offset does not affect PTA likelihoods.
    \label{fig:earthsun}}
\end{figure}

@@

The search for stochastic GW signals with PTAs exploits the distinctive quadrupolar signature in the inter-pulsar correlations of timing-model residuals \citep{hd83}.
Residuals are obtained after applying a chain of corrections that convert the TOA measured at the radio telescope to the notional emission time in the pulsar system (see, e.g., \citealt{ehm06}):
%
\begin{equation}
    t_\mathrm{e}^\mathrm{psr} = t_\mathrm{a}^\mathrm{obs} - \Delta_\odot - \Delta_\mathrm{IS} - \Delta_\mathrm{B},
\end{equation}
%
where $t_\mathrm{e}^\mathrm{psr}$ and $t_\mathrm{a}^\mathrm{obs}$ are the emission and arrival times, $\Delta_\odot$ captures corrections between the observatory and SSB frames, $\Delta_\mathrm{IS}$ describes corrections between the SSB frame and the pulsar-system barycenter, and $\Delta_\mathrm{B}$ models corrections from the pulsar system barycenter to the pulsar frame (which are relevant for binary systems).
Among these terms, $\Delta_\mathrm{IS}$ is very large, but changes very slowly over the duration of pulsar datasets, and thus maps to a constant phase offset. Next comes the R{\o}mer delay \citep{Roemer1676}, corresponding to the light travel time ($\sim 500$ s) between the observatory at $\mathbf{r}^\mathrm{obs}$ and the SSB at $\mathbf{r}^\mathrm{SSB}$:
% MV: the sign here is consistent with the R{\o}mer delay being an actual delay (the pulse takes extra time traveling from SSB-impinging wavefront to observatory)
%
\begin{equation}
\label{eq:Roemer}
\Delta_{\mathrm{R}\odot} =
t_\mathrm{a}^\mathrm{obs} - t_\mathrm{a}^\mathrm{SSB} =
-\frac{\bigl(\mathbf{r}^\mathrm{obs}(t_\mathrm{a}^\mathrm{obs}) - \mathbf{r}^\mathrm{SSB}(t_\mathrm{a}^\mathrm{obs})\bigr) \cdot \hat{\mathbf{p}}}{c},
\end{equation}
%
where $\hat{\mathbf{p}}$ is the unit vector in the direction of the pulsar. In fact, $\hat{\mathbf{p}}$ is determined from pulsar-timing data mainly through the time dependence of Eq.\ \eqref{eq:Roemer}.

@@

\citet{2010ApJ...720L.201C} discuss pulsar-timing's potential to constrain the masses of outer planets. The uncertainties in current IAU best estimates (derived from spacecraft tracking and Doppler studies: \citealt{iaumasses,jh+2000,2006AJ....132.2520J,2014AJ....148...76J,2009AJ....137.4322J}) give rise to R{\o}mer corrections comparable to the orbit errors. These scale as $(\delta m^\mathrm{planet}/m^\mathrm{SS}) \times \mathbf{r}$, so
%
\begin{equation}
    \begin{aligned}
    \delta x^\mathrm{SSB}(\delta m^{(5)}) & \approx (1.6 \!\times\! 10^{-11})\!\times\!
    % (7.8 \!\times\! 10^{8} \, \mathrm{km})
    (5.2 \, \mathrm{AU})
    \approx 12 \, \mathrm{m} \approx 40 \, \mathrm{ns}, \\
    \delta x^\mathrm{SSB}(\delta m^{(6)}) & \approx (8.2 \!\times\! 10^{-12}) \!\times\!
    % (1.4 \!\times\! 10^{9} \, \mathrm{km})
    (9.4 \, \mathrm{AU})
    \approx 12 \, \mathrm{m} \approx 40 \, \mathrm{ns}, \\
    \delta x^\mathrm{SSB}(\delta m^{(7)}) & \approx (3.2 \!\times\! 10^{-11}) \!\times\!
    % (2.9 \!\times\! 10^{9} \, \mathrm{km})
    (19 \, \mathrm{AU})
    \approx 90 \, \mathrm{m} \approx 300 \, \mathrm{ns}, \\
    \delta x^\mathrm{SSB}(\delta m^{(8)}) & \approx (8.0 \!\times\! 10^{-11}) \!\times\!
    % (4.5 \!\times\! 10^{9} \, \mathrm{km})
    (30 \, \mathrm{AU})
    \approx 360 \, \mathrm{m} \approx 1.2 \, \mu\mathrm{s}.
    % (1200 \, \mathrm{ns}).
    \end{aligned}
\end{equation}

These corrections yield delays with the same periods as the corresponding planets, so again they may be observable in PTA datasets for Jupiter and Saturn, but not for Uranus and Neptune.
% MV: say anything about the decay of accuracy past the end of the data?
%
\begin{figure*}[t]
    \centering
    \includegraphics[width=\columnwidth]{roemerdiff.pdf}
    \caption{Differences in R{\o}mer delays computed with DE421/DE430/DE435/DE438 and taking DE436 as reference for 100 simulated pulsars randomly distributed across the sky, plotted after subtracting the best-fitting quadratic and yearly sinusoid. The darker curves with various dashing styles allow a comparison of the same pulsars across the three differences. The plot covers the span of the NANOGrav 11-yr dataset.
    For any given pulsar, these differences are reduced to less than 10 ns after subtracting the best-fitting linear combination of \textsc{BayesEphem} partials, dotted into the pulsar position to obtain R{\o}mer delays.
    }
    \label{fig:Roemer}
\end{figure*}

@@

Note that we do not expect this term to affect GW posteriors: both static and uniform rotations of the SSE frame are absorbed in the estimated positions and proper motions of the pulsars. We omit the former altogether, and include $\omega^{\hat{z}}$ as a check.
%
\item \textbf{TOA delays generated by perturbing Jupiter's average orbital elements}, as given by
\begin{equation}
\label{eq:orbitperturb}
    \delta t_{\mathrm{R}\odot}(\delta a^\mu) = -\delta a^\mu
    \frac{m^{(5)}}{m^\mathrm{SS}} \frac{\partial \mathbf{x}^{(5)}}{\partial a^\mu} \cdot \frac{\mathbf{p}}{c},
\end{equation}
%
where the six $a^\mu$ are the six J2000 Keplerian elements (semi-major axis, eccentricity, inclination, mean longitude, longitude of the perihelion, longitude of the ascending node), and the $\delta a^\mu$ are their perturbations. Alternatively, we can formulate the perturbations in terms of Brouwer and Clemence's ``Set III'' parameters \citep{1961mcm..book.....B}, for which we have access to uncertainty estimates for certain JPL SSEs (\citealt{de434,de438}; see Fig.\ \ref{fig:setIIIposteriors}). The two formulations are largely equivalent with respect to their effect on GW searches.
% quasi-Keplerian parameters
% a: semi-major axis (in AU)
% e: eccentricity (in rad? should be dimensionless)
% I: inclination (in deg)
% L: mean longitude (in deg)
% longitude of the periapsis (in deg)
% longitude of the ascending node (in deg)
% rates in these units per century -> one of the rates corresponds to the period

To implement the R{\o}mer-delay perturbations, we begin with approximate values for the $a^\mu$ and their rates of change.\footnote{The six $a^\mu$ plus the rate of change $\dot{a}^4 \equiv \dot{l}$ of mean longitude specify Jupiter's orbit as the osculating ellipse at the J2000 reference epoch (MJD 2451545); the remaining five rates encode the secular evolution of Jupiter's orbit due to SSB bodies other than the Sun, and to other physical effects. See \href{https://ssd.jpl.nasa.gov/txt/aprx_pos_planets.pdf}{https://ssd.jpl.nasa.gov/txt/aprx\_pos\_planets.pdf}.}
%\citep{ssdelements}.
We vary these $a^\mu$ to minimize the (root-mean-square) difference between our quasi-Keplerian orbits and DE436, integrated between years 2000 and 2020. The resulting orbits are within 1\% of DE436, which ensures similar accuracy for the orbit partials, more than enough for our purposes.
%
\begin{figure*}[t]
    \centering
    \includegraphics[width=\columnwidth]{JupiterPartials.pdf}
    \includegraphics[width=\columnwidth]{JupiterPartialsSVD.pdf}
    \caption{\textbf{Top}: perturbative partials ($x$, $y$, and $z$ coordinates, geometrized units of seconds) for Jupiter's orbit between the years 2000 and 2020, as obtained by varying the six Keplerian elements in a quasi-Keplerian model that includes their rates. The baseline elements and rates are adjusted to fit DE436 orbits.
    Here $a$ is the semi-major axis of the orbit, $e$ the eccentricity, $\imath$ the inclination, $\ell$ the mean longitude, $\varpi$ the longitude of the periapsis, and $\Omega$ the longitude of the ascending node. 
    \textbf{Bottom}: singular-value-decomposition vector time series obtained from the partials in the top panel.}
    \label{fig:jupiterorbit}
\end{figure*}

@@

We do not include Uranus and Neptune's orbital perturbations, which currently lead to $\delta t_{\mathrm{R\odot}}$ with linear time dependencies that are absorbed entirely by timing-model parameters. By contrast, we repeat the procedure that we just outlined for Saturn, but find that GW posteriors are barely affected for orbital perturbations of magnitude comparable to the differences among the JPL SSEs. Thus in our use of \textsc{BayesEphem} we usually omit Saturn orbit perturbations. These may prove more important as datasets increase in length.
\end{itemize}
